### 2017-10-04 abstract.tex
completed first revision of introduction

% 
It has become common practice to validate ground motion simulation based on a variety of time- and frequency-domain metrics known in seismology and engineering and scaled to quantify the level of agreement between synthetics and data or other reference solutions. There is, however, no agreement about the importance or weight that it ought to be given to each individual metric. This leads to their selection often being subjective, either based on intended applications or personal preferences. In turn, this makes it difficult for simulators to identify modeling improvement needs, which could be easier if they could focused on a reduced number of alternative metrics. We present an analysis of ground motion validation metrics using semi-supervised and supervised learning techniques used to label and classify  goodness-of-fit (GOF) results, with the objective prioritizing and narrowing the choice of metrics available and commonly used in ground motion simulation validation. In particular, we study the relationships that exist between eleven different metrics when understood as part of a multi-dimensional space using a constrained \kmeans{} method. We conduct a subspace clustering analysis to address the high-dimensional effects and this allows us to label the results from a large validation dataset into four categories (poor, fair, good, and excellent). We then develop two decision trees using a C5.0 algorithm. These decision trees help narrow the number of metrics leading to a validation prediction into the four categories. These decision trees can be understood as rapid predictors of the quality of a simulation, or as a data-informed classifiers that can help prioritize validation metrics. Our analysis indicates that among the eleven metrics considered here, the acceleration response spectra and total energy of velocity are the most dominant metrics ones, followed by the peak ground response in terms of velocity and acceleration.

% OLD NAEEM VERSION
% 
% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics-based deterministic and hybrid simulation methods. Unlike for low frequency simulations $(f \le 0.5 Hz)$, for which it has become reasonable to expect a good match between synthetics and data, in the case of high-frequency simulations $(1 Hz \le f)$ it is not possible to match results on a wiggle-by-wiggle basis. This is mostly due to the various complexities and uncertainties involved in earthquake ground motion modeling. Therefore, in order to compare synthetics with data we turn to different time series metrics, which are used as means to characterize how the synthetics match the data on qualitative and statistical sense. In general, these metrics provide GOF scores that measure the level of similarity in the time and frequency domains. It is common for these scores to be scaled from 0 to 10, with 10 representing a perfect match. Although using individual metrics for particular applications is considered more adequate, there is no consensus or a unified method to classify the comparison between a set of synthetic and recorded seismograms when the various metrics offer different scores. We study the relationship among these metrics through semi-supervised and supervised learning approaches in order to labeling and classifying data, respectively. In particular, we use constrained \kmeans{} clustering method, in which we define 4 hypothetical stations with scores 3, 5, 7, and 9 for all metrics. We put these stations in the category of cannot-link constraints. We generate the dataset through the validation of the results from a deterministic (physics-based) ground motion simulation for a moderate magnitude earthquake in the greater Los Angeles basin using three velocity models. The maximum frequency of the simulation is 4 Hz. The dataset involves over 300 stations and 11 metrics, or features, as they are understood in the machine learning lingo, where the metrics form a multi-dimensional space. We address the high-dimensional feature effects with a subspace-clustering analysis using 2,3, and 4 dimensional spaces. We label the dataset in each subspace as poor, fair, good, and excellent, where, hypothetical stations with score of 3,5,7, and 9 belong them, respectively. We assign the final label based on the mode of subspaces for each station.  Once the dataset has been labeled, we develop two decision trees using C5.0 algorithm as prediction models to classify a simulation in four groups (poor, fair, good, and excellent) based on a reduced number of metrics. We address the unbalanced dataset issue through oversampling method and analyze the precision, recall and F1-score of the classifiers. The prediction models represent a simple, yet effective model to accurately classify the accuracy of the simulation process based on application-independent approach. We analyze the different combinations of simulation accuracy classes for each components and present a guideline for making decision on accuracy level of simulation independent of components. We apply the proposed method on 2008 Chino Hills earthquake data and synthetic and present the results.

>>>>

% 
It has become common practice to validate ground motion simulation based on a variety of time- and frequency-domain metrics known in seismology and engineering and scaled to quantify the level of agreement between synthetics and data or other reference solutions. There is, however, no agreement about the importance or weight that it ought to be given to each individual metric. This leads to their selection often being subjective, either based on intended applications or personal preferences. In turn, this makes it difficult for simulators to identify modeling improvement needs, which could be easier if they could focused on a reduced number of alternative metrics. We present an analysis of ground motion validation metrics using semi-supervised and supervised learning techniques used to label and classify  goodness-of-fit (GOF) results, with the objective prioritizing and narrowing the choice of metrics available and commonly used in ground motion simulation validation. In particular, we study the relationships that exist between eleven different metrics when understood as part of a multi-dimensional space using a constrained \kmeans{} method. We conduct a subspace clustering analysis to address the high-dimensional effects and this allows us to label the results from a large validation dataset into four categories (poor, fair, good, and excellent). We then develop two decision trees using a C5.0 algorithm. These decision trees help narrow the number of metrics leading to a validation prediction into the four categories. These decision trees can be understood as rapid predictors of the quality of a simulation, or as a data-informed classifiers that can help prioritize validation metrics. Our analysis indicates that among the eleven metrics considered, the acceleration response spectra and total energy of velocity are the most dominant ones, followed by the peak ground response in terms of velocity and acceleration.

% ===========================================================================================
%
% OLD NAEEM VERSION
% 
% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics-based deterministic and hybrid simulation methods. Unlike for low frequency simulations $(f \le 0.5 Hz)$, for which it has become reasonable to expect a good match between synthetics and data, in the case of high-frequency simulations $(1 Hz \le f)$ it is not possible to match results on a wiggle-by-wiggle basis. This is mostly due to the various complexities and uncertainties involved in earthquake ground motion modeling. Therefore, in order to compare synthetics with data we turn to different time series metrics, which are used as means to characterize how the synthetics match the data on qualitative and statistical sense. In general, these metrics provide GOF scores that measure the level of similarity in the time and frequency domains. It is common for these scores to be scaled from 0 to 10, with 10 representing a perfect match. Although using individual metrics for particular applications is considered more adequate, there is no consensus or a unified method to classify the comparison between a set of synthetic and recorded seismograms when the various metrics offer different scores. We study the relationship among these metrics through semi-supervised and supervised learning approaches in order to labeling and classifying data, respectively. In particular, we use constrained \kmeans{} clustering method, in which we define 4 hypothetical stations with scores 3, 5, 7, and 9 for all metrics. We put these stations in the category of cannot-link constraints. We generate the dataset through the validation of the results from a deterministic (physics-based) ground motion simulation for a moderate magnitude earthquake in the greater Los Angeles basin using three velocity models. The maximum frequency of the simulation is 4 Hz. The dataset involves over 300 stations and 11 metrics, or features, as they are understood in the machine learning lingo, where the metrics form a multi-dimensional space. We address the high-dimensional feature effects with a subspace-clustering analysis using 2,3, and 4 dimensional spaces. We label the dataset in each subspace as poor, fair, good, and excellent, where, hypothetical stations with score of 3,5,7, and 9 belong them, respectively. We assign the final label based on the mode of subspaces for each station.  Once the dataset has been labeled, we develop two decision trees using C5.0 algorithm as prediction models to classify a simulation in four groups (poor, fair, good, and excellent) based on a reduced number of metrics. We address the unbalanced dataset issue through oversampling method and analyze the precision, recall and F1-score of the classifiers. The prediction models represent a simple, yet effective model to accurately classify the accuracy of the simulation process based on application-independent approach. We analyze the different combinations of simulation accuracy classes for each components and present a guideline for making decision on accuracy level of simulation independent of components. We apply the proposed method on 2008 Chino Hills earthquake data and synthetic and present the results.

### 2017-10-04 introduction.tex
completed first revision of introduction

Verification and validation of ground motion synthetics have received increasing attention in recent years due to advances in deterministic and non-deterministic physics-based earthquake ground motion simulation. Various methods or schemes have been proposed to evaluate, quantitatively, the misfit or goodness-of-fit (GOF) between simulation synthetics and recorded data or reference alternative solutions \citep{Anderson_2004_Proc, Kristekova_2006_BSSA, Kristekova_2009_GJI, Olsen_2010_SRL, Burks_BSSA_2014, Rezaeian_2015_BSSA}. Among these, there are methods better suited for verification purposes, because they target the comparisons at a waveform level \citep{Kristekova_2006_BSSA, Kristekova_2009_GJI}, whereas others evaluate the comparisons through proxies at the structural response level \citep{Burks_BSSA_2014}. The remaining ones, on the other hand, target specific characteristics in the seismograms themselves in both time and frequency, and are thus better suited for high-frequency or broadband simulations, where the expectation of matching seismograms at the waveform level are less relevant, unless dealing with low-frequency (\fmaxleq{1}). From this group, \citep{Anderson_2004_Proc, Olsen_2010_SRL} are preferred because they offer a simple scoring scheme to normalize the GOF scores associated with different metrics on a common scale.

Out of the various validation methods available, \citet{Anderson_2004_Proc} is perhaps the most widely used \citep[e.g.,]{Chaljub_2010_BSSA, Bielak_2010_GJI, Guidotti_2011_SRL, Maufroy_2015_BSSA}. We, in particular, have consistently relied on a modified version of this method for the validation of a series of simulations of the 2008 $M_w$ 5.4 Chino Hills earthquake, and for the evaluation of velocity models in southern California through the validation of a large suite of moderate-magnitude events \citep{Taborda_2013_BSSA, Taborda_2014_BSSA, Taborda_2016_GJI}. In essence, Anderson's method assess the similarity of two signals based on the average scores of 10 metrics. These metrics measure the cumulative strength of ground motion signals (in terms of Arias intensity and absolute energy), the comparative evolution of the signals in time (through normalized Arias and energy integrals), the peak response (in acceleration, velocity and displacement), the signals' frequency content (in the form of Fourier and response spectra), and their synchronization (through cross-correlation).

While the strength of method of \citet{Anderson_2004_Proc} is that its metrics are well understood in engineering and seismology, others have pointed out that some of these metrics are redundant or incomplete. Both \citet{Taborda_2013_BSSA} and \citet{Maufroy_2015_BSSA}, for instance, added a

To remedy

However, the existence of various alternative methods, including those inspired by or modified after Anderson's criteria, suggest the need to better understand what sort of parameters ought to be taken into account to appropriately classify the validity of a simulation.

Both , for instance, independently suggested the need to add

, and have more recently used it

% OLD NAEEM VERSION

% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics based deterministic and hybrid simulation methods. Unless in very low frequencies (\fleq{0.5}), due to limitations and uncertainties in the ground motion simulation and velocity models, it is not possible to match the simulation and synthetics wiggle by wiggle. Therefor, in order to compare synthetic seismogram with recorded data as a complex time series, different metrics are defined. These metrics are means to characterize the how well the synthetic matches the statistical characteristic of observed data. In general, these metrics compute the goodness-of-fit (GOF) between data and synthetic in time and frequency domains. \citet{Anderson_2004_Proc} proposed 10 metrics to be used to characterize the GOF in different frequency bands. \citet{Kristekova_2006_BSSA,Kristekova_2009_GJI} developed envelope and phase misfit criteria as well as GOF metrics for realistic scenarios. \citet{Olsen_2010_SRL} presented an alternative GOF metrics for broadband synthetics with relatively high resolution of the small misfits, and \citet{Taborda_2013_BSSA} added modification to \citet{Anderson_2004_Proc} metrics.

>>>>

Verification and validation of ground motion synthetics have received increasing attention in recent years due to advances in deterministic and non-deterministic physics-based earthquake ground motion simulation. Various methods or schemes have been proposed to evaluate, through direct quantitative comparisons or indirect statistical analysis, the similarity between simulation synthetics and recorded data or other reference solutions \citep{Anderson_2004_Proc, Kristekova_2006_BSSA, Kristekova_2009_GJI, Olsen_2010_SRL, Burks_BSSA_2014, Rezaeian_2015_BSSA}. Among these methods, there are some that are better suited for verification purposes because they compare the signals at a waveforms level \citep{Kristekova_2006_BSSA, Kristekova_2009_GJI}, whereas others are better suited for application validation purposes because they compare the behavior of systems (e.g., buildings or earth structures) when subjected to alternative input motions through proxies at the response level \citep{Burks_BSSA_2014}. The remaining methods, on the other hand, focus on the overall characteristics of seismograms, in both time and frequency, and are thus better suited for validating high-frequency or broadband simulations, where applications may be undefined and the expectation of matching seismograms at the waveform level is less relevant (unless dealing with low-frequencies, \fmaxleq{1}). From this latter group, methods such as those of \citet{Anderson_2004_Proc} and \citet{Olsen_2010_SRL} are preferred because they offer simple schemes to quantify the goodness-of-fit (GOF) of synthetics with respect to data.

Out of the various validation methods just mentioned, \citet{Anderson_2004_Proc} is perhaps the most widely used today \citep[e.g.,]{Chaljub_2010_BSSA, Bielak_2010_GJI, Guidotti_2011_SRL, Maufroy_2015_BSSA}. We, in particular, have consistently relied on a modified version of this method for the validation of a series of simulations of the 2008 $M_w$ 5.4 Chino Hills earthquake, and for the evaluation of velocity models in southern California through the validation of a large suite of moderate-magnitude earthquake simulations \citep{Taborda_2013_BSSA, Taborda_2014_BSSA, Taborda_2016_GJI}. In essence, Anderson's method assess the similarity of two signals based on the average scores of 10 metrics. These metrics measure the cumulative strength of the signals (in terms of Arias intensity and absolute energy), the comparative evolution of the signals in time (through normalized Arias and energy integrals), the peak values (in acceleration, velocity and displacement), the signals' frequency content (in the form of Fourier and response spectra), and their synchronization (through cross-correlation).

The strength of method proposed by \citet{Anderson_2004_Proc} is that its metrics are well understood in the engineering and seismology contexts. However, it has been pointed out that some of these metrics are redundant, or that other relevant metrics need to be added. \citet{Taborda_2013_BSSA}, for instance, included the strong motion duration \citep{Trifunac_1975_BSSA} as an additional metric, and averaged the Arias intensity- and energy integral-related scores to avoid duplication. In the same spirit, \citet{Maufroy_2015_BSSA} reduced the number of metrics limiting them to only those metrics with comparable units. Unfortunately, none of these alternatives address the underlying questions: what are the parameters that ought to be taken into account when validating ground motion simulations, and what level of priority should these be given.

Lack of consensus about how to answer the above question, makes the choice of validation methods and the selection of the comparison metrics to be a subjective one, based mostly on personal preferences and---arguably---on the application intended for the simulation. In turn, the current multiplicity of metrics, if all are given the same weight in the analysis of validation results, makes it difficult for simulators to identify the sort of changes needed in their models that could lead to better ground motion predictions. We argue that this situation can be corrected if we can offer data-informed arguments they can help simulators focus on a reduced number of alternative metrics.

To that end, we study the relationships that exist between eleven different metrics---those proposed by \citet{Anderson_2004_Proc}, plus strong motion duration---with the objective of offering a prioritized reduced number of metrics that can help predict validation results. We treat these metrics as independent variables defining a multi-dimensional space, and analyze them using machine learning techniques. We use a constrained \kmeans{} method and conduct a subspace clustering analysis to address high-dimensional effects in order to label the results from a large validation dataset into four categories (poor, fair, good, and excellent). The dataset used corresponds to a series of simulations for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake, and their corresponding validation results \citep{Taborda_2014_BSSA}. Upon labeling the dataset, we develop two decision trees using a C5.0 algorithm to prioritize the metrics involved in the validation process. The decision trees narrow the number of metrics offering a prediction algorithm into the aforementioned validation categories. In summary, our results indicate that among the eleven metrics considered in the analysis, the acceleration response spectra and total energy of velocity are the most dominant ones, followed by the peak ground response in terms of velocity and acceleration. We test our prediction model and offer a discussion about its implications and potential use in future validation efforts.

% ===========================================================================================
%
% OLD NAEEM VERSION
% 
% The validation of ground motion synthetics has received increased attention over the last few years due to the advances in physics based deterministic and hybrid simulation methods. Unless in very low frequencies (\fleq{0.5}), due to limitations and uncertainties in the ground motion simulation and velocity models, it is not possible to match the simulation and synthetics wiggle by wiggle. Therefor, in order to compare synthetic seismogram with recorded data as a complex time series, different metrics are defined. These metrics are means to characterize the how well the synthetic matches the statistical characteristic of observed data. In general, these metrics compute the goodness-of-fit (GOF) between data and synthetic in time and frequency domains. \citet{Anderson_2004_Proc} proposed 10 metrics to be used to characterize the GOF in different frequency bands. \citet{Kristekova_2006_BSSA,Kristekova_2009_GJI} developed envelope and phase misfit criteria as well as GOF metrics for realistic scenarios. \citet{Olsen_2010_SRL} presented an alternative GOF metrics for broadband synthetics with relatively high resolution of the small misfits, and \citet{Taborda_2013_BSSA} added modification to \citet{Anderson_2004_Proc} metrics.
